---
title: "SST and Fishing Gear"
author: "Katie O'Dea"
date: "10/25/2021"
output: html_document
---
STEP 1
# load required packages

  # install packages - only need to do this once in the Console
    # install.packages("ncdf4")      ## for reading netCDF files
    # install.packages("rgdal")      ## for r
    # install.packages("sp")    ## for creation of geospatial objects
    # install.packages("sf")
    # install.packages("raster")     ## raster creation and extraction
    # install.packages("maptools")   ## geospatial manipulation
    # install.packages("rgeos")      ## geospatial manipulation
    
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ncdf4)
library(rgdal)
library(sp)
library(sf)
library(raster)
library(maptools)
library(rgeos)

```

#STEP 2
#Pulling out the dimensions of the data
```{r import data}
# importing the gear data from SAU
raw_gear_data <- read_csv("data/SAU EEZ 958,959,851,852,848,850 v48-0.csv")

# opening the nc file	from NOAA
raw_sst_data <- nc_open("data/sst.mnmean.nc")
#checking attributes
print(raw_sst_data) # 2 variables, excluding dimension variables. 
    # Monthly Mean of Sea Surface Temp in degrees Celsius
    # Missing data is flagged with a value of 32767s
    # 4 dimensions are longitude (units are degrees east; size 360), latitude (units are degrees north; size 180), time (units are days since 1800-1-1; size 478), and nbnds (unclear what this is?).
    # time: since observations are monthly means from 1981/12 to 2021/09, it makes sense there are 478 observations (although may be missing a few observations somewhere)
    # 10 global attributes - not sure what these mean

```

#extracting variables 
```{r extract variables}
lon <- ncvar_get(raw_sst_data,"lon")
nlon <- dim(lon) # 360
head(lon) #0.5 1.5 2.5 3.5 4.5 5.5
    
lat <- ncvar_get(raw_sst_data,"lat")
nlat <- dim(lat) # 180
head(lat) #89.5 88.5 87.5 86.5 85.5 84.5

print(c(nlon,nlat)) #360 180

time <- ncvar_get(raw_sst_data,"time")
head(time) #66443 66474 66505 66533 66564 66594 66625 66655 66686

tunits <- ncatt_get(raw_sst_data,"time","units")
    nt <- dim(time)
    nt #478

lonlat <- as.matrix(expand.grid(lon,lat)) # making a grid and then creating a matrix
    dim(lonlat) # 1129600 obs and 2 vars? (I think?)
    str(lonlat)
    
# get temperature
    sst_array <- ncvar_get(raw_sst_data,"sst")
    sstlongname <- ncatt_get(raw_sst_data,"sst","long_name")#Monthly Mean of Sea Surface Temperature 
    sstunits <- ncatt_get(raw_sst_data,"sst","units") #degrees Celsius
    dim(sst_array) # 360 180 478
# replace netCDF fill values with NA's
    sst_array[sst_array==32767] <- NA
    
# close netcdf data - I have extracted everything that I need
    nc_close(raw_sst_data)
  
# delete the associate R data frame to save space
    rm(raw_sst_data)
    
```

# STEP 3
# make the grid points from the data into a spatial dataframe
```{r make grid points into spatial dataframe}

# making a grid_id variable 
    gridsize <-nlat*nlon
    grid_id <- c(1:gridsize) 
    head(grid_id)
    str(grid_id)

# binding the lonlat grid and grid_ids 
    era1_grid_points<-cbind(lonlat,grid_id) 
  
# giving column names
    colnames(era1_grid_points) <-c("longitude", "latitude", "grid_id")  
    str(era1_grid_points)
    head(era1_grid_points) # Looks good 
  
# making the matrix into a dataframe
    era1_grid_pointsdf <- as.data.frame(era1_grid_points) 
  
# convert era1_grid_pointsdf into a sp object
   coordinates(era1_grid_pointsdf) <- ~longitude + latitude # longitude first
  
# inspecting it
   str(era1_grid_pointsdf) # data component is labeled grid_id, looks good
  
# add a coordinate reference system (CRS).
   proj4string(era1_grid_pointsdf) <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +units=km")
  
# inspecting it
    str(era1_grid_pointsdf) 
#need shape files for EEZ
```

# STEP 4
# convert sea surface temperature from netcdf format to SpatialPointsDataFrame format
```{r convert sst}
# check dimensions of temperature 
    dim(sst_array) # 360 by 180 by 478
    
# track the number of time steps - subset of data (don't think I actually need to do this)

# make sst array into vector
    tmp_vec <- as.vector(sst_array)  
    length(tmp_vec) # 360*180*478 = 30974400
    
# delete sst_array to save space
    rm(sst_array)

# reshape vector into a 180*360=64800 by 478 matrix using the matrix() function, verify dimensions & inspect
    sst.mat <- matrix(tmp_vec,  nrow = nlon * nlat, ncol = nt) 
    dim(sst.mat) #  64800 478
    head(sst.mat[,1:20]) # a grid of temperature values (is this correct?)

# label columns of matrix 
    colnames(sst.mat) <- colnames(sst.mat, do.NULL = FALSE, prefix = "tmp")

# add grid_id column to tmp.mat 
    head(grid_id)
    sst.mat <- cbind(grid_id,sst.mat) 
    head(sst.mat[,1:20])

# merge tmp.mat with grid points to make a spatial data frame 
    sstGridPtsdf <- merge(era1_grid_pointsdf, sst.mat, by.x = "grid_id", by.y = "grid_id")  
    str(sstGridPtsdf) # TBD obs. of  1461 variables 
    
#check points by checking grid points and degrees 
    
```

